A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-09-13 submission/manuscript.rmd
more changes for response

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{8mm}

@@

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of propensity scores within strata of predicted risk and
estimation of relative and absolute treatment effect within strata of predicted
risk; 5) evaluation and presentation of results. We demonstrate our framework by
evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
inhibitors versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of
heterogeneity of treatment effect on observational data that are mapped to the
OMOP Common Data Model. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

@@

\> > EDIT > >

***In recent years, a large number of methods has been developed for the
assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
HTE analyses into exploratory, confirmatory, descriptive and predictive
[@Varadhan2013]. Exploratory analyses focus on hypothesis generation,
confirmatory analyses test subgroup effect hypotheses, descriptive analyses aim
at facilitating future synthesis of subgroup effects and predictive analyses
predict probabilities of benefit or harm in individual patients. Predictive HTE
approaches can be further subdivided into risk modeling, treatment effect
modeling and optimal treatment regime methods, based on the reference class used
for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. We focus on “risk modeling” approaches where
patients are divided into risk strata using either an existing or an internally
developed risk prediction model. Risk-stratum-specific estimates provide an
overview of the evolution of treatment effects with increasing risk both on the
relative and the absolute scale. Recently, systematic guidance on the
application of such methods has been developed [@Kent2010; @Kent2018].***

< < EDIT < <

\> > OLD > >

Predictive HTE analyses have been described as approaches that
provide predictions of potential outcomes in a particular patient with one
intervention versus an alternative, taking into account multiple relevant
patient characteristics [@Varadhan2013]. We focus on “risk modeling”
approaches where patients are divided into risk strata using either an existing
or an internally developed risk prediction model. Risk-stratum-specific
estimates provide an overview of the evolution of treatment effects with
increasing risk both on the relative and the absolute scale. Recently,
systematic guidance on the application of such methods has been developed
[@Kent2010; @Kent2018].

< < OLD < <

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding, while the unique structure of
different databases calls for database-specific analysis plans that often are
not easily transportable.

\> > ADD > >

***Because of the latter issue, running analyses at scale demands a big investment
of time and effort, as researchers are forced to map their analysis plans to the
databases available to them.***

< < ADD < <

@@

\> > EDIT > >

***We aimed to develop a framework for implementing risk-based assessment of
treatment heterogeneity in high-dimensional observational data, extending the
existing guidelines of the RCT setting. Our publicly available package provides
an out-of-the-box solution for implementing such analyses at scale within the
OHDSI network, taking advantage of the OMOP-CDM.***

< < EDIT < <

\> > OLD > >

We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity
in high-dimensional observational data, which extends the existing guidelines of
the RCT setting.

< < OLD < <

We implemented the framework using existing OHDSI methods including the
patient-level prediction framework and the population-level effect estimation
framework based on new-user cohort design [@Reps2018; @Ryan2013]. As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. We developed an open-source
R-package for the implementation of the proposed framework and made it publicly
available. The source code can be found at
<https://github.com/OHDSI/RiskStratifiedEstimation>. An overview of the
procedure can be seen in Figure \@ref(fig:figure1).

@@

\> > ADD > >

***Cohort definitions are crucial for this step of the framework. We define a
cohort as the set of patients who satisfy one or more inclusion criteria for a
duration of time. A cohort within the OHDSI setting is more than a set of
specific clinical codes, providing a definition of a logic for how to use that
code set. All cohort definitions consist of: an entry event, i.e. the time a
patient enters a cohort; a set of inclusion criteria applied to the initial
event cohort to further restrict the set of people, resulting in the
construction of the construction of the qualifying cohort; cohort exit criteria
that terminate the patient's presence in the cohort. Cohort definitions are
transportable, meaning that in theory they can be implemented in any database,
provided that it is mapped to the OMOP-CDM.***

< < ADD < <

@@

The aim of this step is the inclusion of databases that represent the patient
population of interest. It is required that the databases are mapped to the
OMOP-CDM. The inclusion of multiple databases potentially increases the
generalizability of results. Furthermore, the cohorts should preferably have
adequate sample size to ensure precise effect estimation, even within smaller
risk strata.

@@

< < MOVE: ps_matching > >

\> > ADD > >

***Other methods of fitting the propensity scores, such as random forest and
others (REF) can also be considered.***

< < ADD < <

@@

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Any appropriate method
for the evaluation of relative and absolute treatment effects can be considered,
as long as the this is done consistently in all risk strata. Common approaches
are odds ratios or hazard ratios for relative scale estimates and differences in
observed proportions or differences in Kaplan-Meier estimates for absolute scale
estimates, depending on the problem at hand. We estimate propensity scores
within risk strata which we then use to match patients from different treatment
cohorts or stratify them into groups with similar propensity scores or to weigh
each patient's contribution to the estimation process [@Austin2011].

@@

@@

\> > MOVE: ps_matching > >

***More specifically, we first estimate propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices. We
match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
for any match. The default value we use is 0.2 on the standardized logit scale
for the propensity scores.***

< < MOVE < <

@@

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are usually adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Multiple outcome risk
stratification schemes can also be considered. However, this should be done with
caution, as it may hinder the interpretability of the results, in a similar
manner as typical subgroup analyses.

@@

\> > EDIT > >

***Ideally, externally derived and adequately validated prediction model would
be preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
absence of such prediction models an internally-developed risk prediction model
can be considered. Earlier simulations of RCT studies have shown that internal
models developed on the combined treatment and control arms blinded to treatment
gave relatively unbiased estimates of treatment effect across the spectrum of
risk [@Burke2014].***

< < EDIT < <

\> > OLD > >

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014].

< < OLD < <

However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, spurious treatment-covariate
interactions may be introduced in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{8mm}

@@

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of propensity scores within strata of predicted risk and
estimation of relative and absolute treatment effect within strata of predicted
risk; 5) evaluation and presentation of results. We demonstrate our framework by
evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
inhibitors versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.05% to -0.02% in the lowest risk quarter to 
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of
heterogeneity of treatment effect on observational data that are mapped to the
OMOP Common Data Model. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

@@

In recent years, a large number of methods has been developed for the
assessment of HTE, mainly in the RCT setting. Earlier work suggested separating
HTE analyses into exploratory, confirmatory, descriptive and predictive
[@Varadhan2013]. Exploratory analyses focus on hypothesis generation,
confirmatory analyses test subgroup effect hypotheses, descriptive analyses aim
at facilitating future synthesis of subgroup effects and predictive analyses
predict probabilities of benefit or harm in individual patients. Predictive HTE
approaches can be further subdivided into risk modeling, treatment effect
modeling and optimal treatment regime methods, based on the reference class used
for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. We focus on “risk modeling” approaches where
patients are divided into risk strata using either an existing or an internally
developed risk prediction model. Risk-stratum-specific estimates provide an
overview of the evolution of treatment effects with increasing risk both on the
relative and the absolute scale. Recently, systematic guidance on the
application of such methods has been developed [@Kent2010; @Kent2018].

While these approaches were developed for application in randomized controlled
trials (RCTs), observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, including heterogeneous populations. However, unlike RCTs,
treatment effects are subject to confounding, while the unique structure of
different databases calls for database-specific analysis plans that often are
not easily transportable. Because of the latter issue, running analyses at scale
demands a big investment of time and effort, as researchers are forced to map
their analysis plans to the databases available to them.

@@

We aimed to develop a framework for implementing a risk-based predictive
approach for evaluating HTE in high-dimensional observational data, extending
the existing guidelines of the RCT setting. Our publicly available package
provides an out-of-the-box solution for implementing such analyses at scale
within the OHDSI network, taking advantage of the OMOP-CDM. We implemented the
framework using existing OHDSI methods including the patient-level prediction
framework and the population-level effect estimation framework based on new-user
cohort design [@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed
heterogeneity of the effects of first-line hypertension treatment: we compared
the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers
on 9 outcomes across three different US claims databases.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation of the results. We developed an open-source R-package for the
implementation of the proposed framework and made it publicly available. The
source code can be found at <https://github.com/OHDSI/RiskStratifiedEstimation>.

@@

Cohort definitions are crucial for this step of the framework. We define a
cohort as the set of patients who satisfy one or more inclusion criteria for a
duration of time. A cohort within the OHDSI setting is more than a set of
specific clinical codes, providing a definition of a logic for how to use that
code set. All cohort definitions consist of: an entry event, i.e. the time a
patient enters a cohort; a set of inclusion criteria applied to the initial
event cohort to further restrict the set of people, resulting in the
construction of the construction of the qualifying cohort; cohort exit criteria
that terminate the patient's presence in the cohort. Cohort definitions are
transportable, meaning that in theory they can be implemented in any database,
provided that it is mapped to the OMOP-CDM.

Our framework uses a comparative cohort design. This means that at least 3
cohorts of patients need to be defined at this stage of the framework.

@@

The aim of this step is the inclusion of databases that represent the patient
population of interest. It is required that the databases are mapped to the
OMOP-CDM. The inclusion of multiple databases potentially increases the
generalizability of results. Furthermore, the cohorts should preferably have
adequate sample size with adequate follow-up time to ensure precise effect
estimation, even within smaller risk strata. Other issues that may be of
importance for database inclusion are the depth of data capture (the precision
at which measurements, lab tests, conditions are recorded), the reliability of
data entry and many more, also depending on the task at hand.

@@

More specifically, we first estimate propensity scores using LASSO logistic
regression and a large set of baseline covariates including demographics, drug
exposures, diagnoses, measurements and medical devices. We match patients 1-1
using a caliper, i.e. the maximum distance that is acceptable for any match. The
default value we use is 0.2 on the standardized logit scale for the propensity
scores. Other methods of fitting the propensity scores, such as random forest
and others can also be considered.

@@

The aim of this step is the estimation of treatment effects (both on the
relative and the absolute scale) within risk strata---typically 4 risk
quarters---defined using the prediction model of step 3. Effect estimation may
be focused on the difference in outcomes for a randomly selected person from the
risk stratum (average treatment effect) or for a randomly selected person from
the treatment cohort within the risk stratum receiving the treatment under study
(average treatment effect on the treated).

Any appropriate method for the evaluation of relative and absolute treatment
effects can be considered, as long as the this is done consistently in all risk
strata. Common approaches are odds ratios or hazard ratios for relative scale
estimates and differences in observed proportions or differences in Kaplan-Meier
estimates for absolute scale estimates, depending on the problem at hand. We
estimate propensity scores within risk strata which we then use to match
patients from different treatment cohorts or stratify them into groups with
similar propensity scores or to weigh each patient's contribution to the
estimation process [@Austin2011].

@@

A schematic overview of the prediction and estimation steps is presented in
Figure \@ref(fig:figure1).

@@

@@

Our aim was to estimate the average treatment effects on the relative and the
absolute scale within strata of predicted acute MI risk.

@@

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are usually adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. It is an addition to the rapidly
expanding literature of approaches for evaluating treatment effect
heterogeneity. Multiple outcomes can be evaluated in patient subgroups of
similar baseline outcome risk. Multiple outcome risk stratification schemes can
also be considered. However, this should be done with caution, as it may hinder
the interpretability of the results, in a similar manner as typical subgroup
analyses.

@@

Ideally, externally derived and adequately validated prediction model would be
preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
absence of such prediction models an internally-developed risk prediction model
can be considered. Earlier simulations of RCT studies have shown that internal
models developed on the combined treatment and control arms blinded to treatment
gave relatively unbiased estimates of treatment effect across the spectrum of
risk [@Burke2014]. However, in observational databases treatment arms may
significantly differ in sample size. Because the prediction model will possibly
better fit to the larger treatment arm, spurious treatment-covariate
interactions may be introduced in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.
